
A huge thank you to my advisor, Vivek Srikumar, for graciously being willing to advise me despite the fact that I read the floating point standard for fun. I couldn't have had a better advisor; Vivek is a master peddler of ideas and an unrelenting optimist-- he consistently convinced me that the research we were working on was interesting and important whenever I lost faith. I really appreciate both being given space to explore my research interests as well as direction and experience working on projects aligned with my interests, even while they were somewhat outside of his. In addition to learning at least a little about how research is done (it's at the twelfth hour), I hope I've learned enough from Vivek about story telling, patience and confidence to apply these practices in the future.

Thank you to everyone involved with SweetPea: Matthew Flatt, Jonathan Cohen, Sebastian Musslick and Ben Draut. Thanks to Jon \& Sebastian for so enthusiastically providing the motivation for this project; it's much more fun to work on something that someone really actually wants to use. Thanks to Matt for his guidance and his patience as I discovered details he saw all along, and to Ben for taking the helm to get SweetPea into the wild.

Thanks to the incredibly patient crew who proof-read everything I've written in the past few years- Tobin Yehle, Dasha Pruss, Nic Bertagnolli \& Zach Price. Thank you to Will Byrd for making me upset with the state of computing and to Michael Adams for telling me how all my academic interests were related.

Thanks also to my friends from the Recurse Center for many delightful computer science distractions, and to my friends in SLC for distractions of other flavors. Thank you to Tobin, Nic and Laura for always being on the lookout for the next scheme, wild idea, and adventure, and for inviting me along-- and sometimes carrying my pack.

And, of course, thank you to my parents-- I'm sorry I didn't become a mathematician-- and to my sister-- I'm sorry I didn't become an artist.

Life is rolling on, and it's all very exciting.
