%%% -*-LaTeX-*-

\chapter{Motivation}
 

The replicability crisis in experimental science is fueled by a lack of transparent and explicit discussion of experimental design in published work; a scientific conclusion is only as trustworthy as the experimental design on which it is based. Incorrectly designed or biased experiments lead to possibly invalid or biased conclusions; therefore creating correct, statistically unbiased, reproducible experimental designs is paramount to performing meaningful experiments that produce trustworthy and generalizable results.

This thesis describes \textbf{SweetPea}, a system for describing experimental designs and generating experimental sequences specified by the design in a statistically rigorous way. SweetPea's high-level language allows scientists to declaratively describe the experiment they want to conduct rather than forcing them to procedurally describe how to construct their experimental sequences. This allows scientists to write concise, correct programs which describe their experiments, and produce unbiased experimental sequences; these programs can then be published and shared to document the experiment and facilitate interpretation and replication.

\section{Reliable Experimental Design}

What does experimental design entail? A single experiment is a sequence of trials, each of which is characterized by several factors. When the experiment is run, certain factors, such as response time, may be aggregated to produce statistical evidence to validate or disprove a hypothesis. A design is a specification of the trials and any invariants that need to hold for the experiment; the design of an experiment specifies which sequences of trials are legal to include for that experiment.

While there are many software tools for modeling and running experiments (e.g., \cite{cohen1993psyscope} \cite{mathot2012opensesame} \cite{peirce2009generating}), there are, to the best of our knowledge, none for designing them. To ensure statistically interpretable results, every possible trial sequence that satisfies the constraints of the experimental design must have an equal likelihood of being chosen for the experiment. This guarantees that the method for generating trial sequences is not introducing bias. In practice, however, researchers often construct trial sequences without statistical guarantees. For even moderately complex designs, the number of valid sequences can quickly become both intractably large and sparse in the space of all sequences, making it difficult or even impossible to find a valid sequence by randomly sampling all sequences or by enumerating all valid sequences.

Currently, scientists design experiments by writing complex scripts which manually balance the experimental factors of interests. There are two major issues with this approach: the first is that it may not (and often does not) produce unbiased sequences of trials. In practice, researchers construct these sequences without any statistical guarantees because the brute force solution for constructing unbiased sequences by enumerating all options is intractable; for a typical experiment the size of the search space can be as large as $10^{100}$. The second issue is that this approach is brittle. It is easy to introduce bugs that go unnoticed but which may have serious consequences, and it is difficult to verify and reproduce another researcher's implementation.

SweetPea can be viewed as a domain-specific interface to SAT-sampling, and while there are other languages that rely on SAT-solvers \cite{torlak2014lightweight}, none that we know of leverage the guarantees provided by SAT-samplers. SweetPea generates unbiased sequences of trials given satisfiable constraints. At the heart of the bias problem is the need to sample from constrained combinatorial spaces with statistical guarantees; SweetPea samples sequences of trials by compiling experimental designs into boolean logic, which are then passed to a SAT-sampler. The SAT-sampler Unigen \cite{meel2016constrained}
provides statistical guarantees that the solutions it finds are approximately uniformly probable in the space of all valid solutions. This means that while producing sequences of trials that are perfectly unbiased is intractable, we do the next best thing-- produce sequences that are guaranteed to be \emph{approximately} unbiased.

\section{Declarative Programming}

Virtually all fields of scientific research increasingly rely on computational tools. Computational tools open the door to analyses that are otherwise intractable, time consuming and error-prone. Moreover, writing programs contributes to making research reproducible because it creates digital artifacts, which allow one scientist to share, analyze and run a program written by another. The drawback, however, is that writing correct, maintainable, complex programs takes significant engineering effort. Requiring scientists to be engineers in addition to being highly trained domain specialists needlessly impedes the progress of science. Declarative languages allow their users to describe the result they want, as contrasted with imperative or procedural languages which require their users to describe how to construct the result.

There is a need, then, for software systems that allow scientists within a given domain to design unbiased, replicable experiments in that domain. Moreover, such systems need to provide an easy-to-use, declarative interface so that scientists who are not necessarily trained as software engineers can create and reason about complicated experimental designs, and transparently share their experimental setups and design choices. SweetPea is just such a system; it is a language which provides semantics for describing sequential trial-based experiments, and a runtime for synthesizing experimental sequences from specifications. While the need for a system to automate experimental design is general to many types of science, we have built a prototype targeted for cognitive psychology and neuroscience, in which trial-based designs are common and issues of reproducibility and complexity of design have become a focus of attention. We discuss what would needed to be added to our prototype to support other experimental sciences in the Future Work section.

SweetPea is a vision of what this needed system could be. This thesis documents the motivation, goals, and current state and future vision for SweetPea.



%%% Index phrases should be attached to an important word of a phrase,
%%% and are usually best kept on a separate line by terminating the
%%% previous line with a percent comment without intervening space, as
%%% in this example:
%%%
%%%     \newcommand {\X} [1] {#1\index{#1}}
%%%
%%%     African ungulates,%
%%%     \index{African ungulate}
%%%     like the \X{gnu}, \X{impala}, \X{kudu}, and \X{springbok}
%%%     live mostly in hot climate and consume vegetation.
%%%
%%% However, for this document, we only want lots of index entries to
%%% populate a sample topic index.

%%% ====================================================================
%%% Cross-references for index entries should be specified only once:
